---
title: "Public Computing Statistics"
format:
  html:
    theme: cosmo
    df-print: kable
    toc: true
    toc-expand: true
---

Script to read an Excel spreadsheet produced by Cybrarian and do some analysis.  **Limitation:** dates of exam week are hard-coded below.

```{r message=FALSE}
library(tidyverse)
library(readxl)

exam_week_start <- date("2024-06-08")
exam_week_end <- date("2024-06-14")

in_exam_week <- function(timestamp) {
  between(date(timestamp), exam_week_start, exam_week_end)
}
```

# Load data

Expected columns and sample row:

Item | Date | Login Time | Computer | User ID | User Name | Session Minutes
---- | ---- | ---------- | -------- | ------- | --------- | ---------------
1 | Monday April 1, 2024 | 7:57:09 AM | PC209-A1HG | warren390 | warren390 | 2

```{r warning=FALSE, message=FALSE}
# Eliminate first 6 and last 5 rows
data <- read_xlsx("data/input.xlsx", skip=6)
data <- read_xlsx("data/input.xlsx", skip=6, n_max=nrow(data)-5)

# Incorporate the date column into the time column to form a unified
# timestamp column
date(data$`Login Time`) <- mdy(str_remove(data$Date, "^\\w+ "))

# Select and rename columns of interest
data <- select(data,
  login_time=`Login Time`,
  computer=Computer,
  session_minutes=`Session Minutes`
)

# The next processing follows the instructions given in:
# https://ucsb-atlas.atlassian.net/wiki/spaces/LIB/pages/17314414653/Public+Computing+Annual+Statistics

# Remove virtual machines
data <- filter(data, !str_starts(computer, "PSY|PT|SYS"))

# Map computers to buildings and floors.  N.B.: code farther down
# relies on how building names are formatted here.
data <- mutate(data,
  building=case_when(
    str_starts(computer, "OAM"     ) ~ "Music",
    str_starts(computer, "OA[^M]"  ) ~ "B (middle 2-story)",
    str_starts(computer, "PA"      ) ~ "A (north 3-story)",
    str_starts(computer, "PB"      ) ~ "B (middle 2-story)",
    str_starts(computer, "PC[124]" ) ~ "D (south 4-story)",
    str_starts(computer, "PC[5678]") ~ "C (tower 8-story)",
    str_starts(computer, "PFM"     ) ~ "Music",
    str_starts(computer, "PF[^M]"  ) ~ "A (north 3-story)",
    str_starts(computer, "PM"      ) ~ "Music",
    str_starts(computer, "PS2"     ) ~ "C (tower 8-story)",
    str_starts(computer, "PS3"     ) ~ "A (north 3-story)", # Wiki says "SRC"
    str_starts(computer, "PW"      ) ~ "A (north 3-story)",
    str_starts(computer, "RSV"     ) ~ "B (middle 2-story)",
    TRUE ~ NA
  ),
  floor=case_when(
    str_starts(computer, "OAM|PFM|PM") ~ 2,
    str_starts(computer, "RSV") ~ 1,
    str_starts(computer, "^[A-Z]+\\d\\d\\d-") ~
      as.integer(sub("^[A-Z]+(\\d).*", "\\1", computer)),
    TRUE ~ NA
  )
)

# For certain plots it will be convenient to be able to refer to a
# building and floor more compactly as an "area"
data <- mutate(data,
  area=paste(str_sub(building, 1, 1), floor, sep="")
)

# Fix up datatypes
data <- mutate(data,
  computer=as.factor(computer),
  session_minutes=as.integer(session_minutes),
  building=as.factor(building),
  floor=as.integer(floor),
  area=as.factor(area)
)

# Create a dataframe listing the areas, for use in joins
areas <- unique(select(data, building, floor, area)) %>%
  arrange(building, floor)

# Number of computers in each area (inferred from data, not ground
# truth)...
computers <- unique(select(data, building, floor, area, computer)) %>%
  group_by(building, floor, area) %>%
  summarize(num_computers=n())

# ...and in each building
computers_by_building <- computers %>%
  group_by(building) %>%
  summarize(num_computers=sum(num_computers))

# For convenience, create a dataframe that counts numbers of logins
login_data <- data %>%
  group_by(date=date(login_time), building, floor, area) %>%
  summarize(num_logins=n())
```

# Analysis

## Logins

### Average logins per day

```{r}
#| code-fold: true
df <- login_data %>% group_by(date) %>% summarize(num_logins=sum(num_logins))
round(mean(df$num_logins), 0)
```

### Average logins per day, by area

```{r message=FALSE}
#| code-fold: true
login_data %>%
  group_by(Building=building, Floor=floor) %>%
  summarize(`Average daily logins`=round(mean(num_logins), 0))
```

## Duration

### Average session duration

In minutes.

```{r}
#| code-fold: true
round(mean(data$session_minutes), 0)
```

### Average session duration, by area

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(Building=building, Floor=floor) %>%
  summarize(`Average session duration (min)`=round(mean(session_minutes), 0))
```

## Occupancy

Here we consider the percentage of time the computers in an area are occupied.  An area has an occupancy of 100% if every computer in the area is occupied by a session every minute the Library is open.  Two notes:

- The number of computers in each area is inferred from those mentioned in the Cybrarian input data; this notebook has no independent source of information about numbers of computers.

- For the purpose of calculating percentages, the Library is assumed to be open 12 hours per day.

```{r}
open_minutes <- 12*60 # half a day
```

### Average daily occupancy

Over all areas of the Library, as a percentage of theoretical capacity (`r sum(computers$num_computers)` computers total, available `r open_minutes` minutes per day).

```{r message=FALSE}
#| code-fold: true
df <- data %>%
  group_by(date(login_time)) %>%
  summarize(minutes_used=sum(session_minutes))

round(mean(df$minutes_used)/(sum(computers$num_computers)*open_minutes)*100, 0)
```

### Average daily occupancy, by area

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(date(login_time), area) %>%
  summarize(minutes_used=sum(session_minutes)) %>%
  group_by(area) %>%
  summarize(avg_minutes_used=mean(minutes_used)) %>%
  inner_join(computers, by=join_by(area)) %>%
  mutate(
    occupied_pct=round(avg_minutes_used/(num_computers*open_minutes)*100, 0)
  ) %>%
  select(Building=building, Floor=floor, `% occupied`=occupied_pct)
```

## Effects of exam week

Comparing usage during exam week and not.

### Average logins per day

```{r message=FALSE}
#| code-fold: true
login_data %>%
  group_by(date) %>%
  summarize(num_logins=sum(num_logins)) %>%
  group_by(`Exam week?`=in_exam_week(date)) %>%
  summarize(`Average daily logins`=round(mean(num_logins), 0))
```

### Average logins per day, by building and area

```{r message=FALSE}
#| code-fold: true
login_data %>%
  group_by(date, building) %>%
  summarize(num_logins=sum(num_logins)) %>%
  group_by(`Exam week?`=in_exam_week(date), Building=building) %>%
  summarize(`Average daily logins`=mean(num_logins)) %>%
  ggplot(aes(x=Building, y=`Average daily logins`, fill=`Exam week?`)) +
    geom_bar(stat="identity", position="dodge")
```

```{r message=FALSE}
#| code-fold: true
login_data %>%
  group_by(
    `Exam week?`=in_exam_week(date),
    `Area (= building + floor)`=area
  ) %>%
  summarize(`Average daily logins`=mean(num_logins)) %>%
  ggplot(
      aes(
        x=`Area (= building + floor)`,
        y=`Average daily logins`,
        fill=`Exam week?`
      )
    ) +
    geom_bar(stat="identity", position="dodge")
```

### Average session duration

```{r}
#| code-fold: true
data %>%
  group_by(`Exam week?`=in_exam_week(login_time)) %>%
  summarize(`Average session duration (min)`=round(mean(session_minutes), 0))
```

### Average session duration, by building and area

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(`Exam week?`=in_exam_week(login_time), Building=building) %>%
  summarize(`Average session duration (min)`=mean(session_minutes)) %>%
  ggplot(
      aes(x=Building, y=`Average session duration (min)`, fill=`Exam week?`)
    ) +
    geom_bar(stat="identity", position="dodge")
```

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(
    `Exam week?`=in_exam_week(login_time),
    `Area (= building + floor)`=area
  ) %>%
  summarize(`Average session duration (min)`=mean(session_minutes)) %>%
  ggplot(
      aes(
        x=`Area (= building + floor)`,
        y=`Average session duration (min)`,
        fill=`Exam week?`
      )
    ) +
    geom_bar(stat="identity", position="dodge")
```

### Average daily occupancy

Over all areas of the Library, as a percentage of theoretical capacity (`r sum(computers$num_computers)` computers total, available `r open_minutes` minutes per day).

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(`Exam week?`=in_exam_week(login_time), date(login_time)) %>%
  summarize(minutes_used=sum(session_minutes)) %>%
  group_by(`Exam week?`) %>%
  summarize(
    `% occupied`=round(
      mean(minutes_used)/(sum(computers$num_computers)*open_minutes)*100,
      0
    )
  )
```

### Average daily occupancy, by building and area

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(
    `Exam week?`=in_exam_week(login_time),
    building,
    date(login_time)
  ) %>%
  summarize(minutes_used=sum(session_minutes)) %>%
  group_by(`Exam week?`, Building=building) %>%
  summarize(avg_minutes_used=mean(minutes_used)) %>%
  inner_join(computers_by_building, by=join_by(Building==building)) %>%
  mutate(
    `% occupied`=round(avg_minutes_used/(num_computers*open_minutes)*100, 0)
  ) %>%
  ggplot(aes(x=Building, y=`% occupied`, fill=`Exam week?`)) +
    geom_bar(stat="identity", position="dodge")
```

```{r message=FALSE}
#| code-fold: true
data %>%
  group_by(`Exam week?`=in_exam_week(login_time), area, date(login_time)) %>%
  summarize(minutes_used=sum(session_minutes)) %>%
  group_by(`Exam week?`, `Area (= building + floor)`=area) %>%
  summarize(avg_minutes_used=mean(minutes_used)) %>%
  inner_join(computers, by=join_by(`Area (= building + floor)`==area)) %>%
  mutate(
    `% occupied`=round(avg_minutes_used/(num_computers*open_minutes)*100, 0)
  ) %>%
  ggplot(
      aes(x=`Area (= building + floor)`, y=`% occupied`, fill=`Exam week?`)
    ) +
    geom_bar(stat="identity", position="dodge")
```

## Peak usage

Cybrarian provides the start and end times of each session.  Armed with this data, we can compute how many computers are in **simultaneous** use.  The analysis below groups computers by building and floor, but does not otherwise distinguish between computers (i.e., all computers within an area are considered equivalent).  First, prep the data...

```{r}
# The loop below takes a looong time, hence the cache file.  While
# there are more efficient algorithmic approaches, it's surprising how
# slow R is...

cache_file <- "data/in_use_data.RData"

if (file.exists(cache_file)) {
  load(cache_file)
} else {

  # Create a dataframe with a row for each minute and each area.
  # Subtle point: we are not capturing the decay in usage that extends
  # beyond the minute of the last login time.
  earliest_mins <- as.integer(min(data$login_time)) %/% 60
  latest_mins <- as.integer(max(data$login_time)) %/% 60
  in_use_data <- crossing(
    timestamp=as_datetime((earliest_mins:latest_mins)*60),
    areas
  )
  in_use_data$num_in_use <- 0
  in_use_data <- mutate(in_use_data,
    num_in_use=as.integer(num_in_use)
  )

  for (i in rownames(data)) { # For each login/session
    start <- data[i,]$login_time
    second(start) <- 0
    end <- start + minutes(data[i,]$session_minutes-1)
    rows <- between(in_use_data$timestamp, start, end) &
      in_use_data$building == data[i,]$building &
      in_use_data$floor == data[i,]$floor
    in_use_data[rows,"num_in_use"] <- in_use_data[rows,"num_in_use"] + 1
  }

  save(in_use_data, file=cache_file)

}
```

### Overall maximums, by area

Again, this is counting computers that are in simultaneous use.  And the comment under [Occupancy](#occupancy) above applies here as well: the number of computers in each area is inferred from those mentioned in the Cybrarian input data.

```{r message=FALSE}
#| code-fold: true
in_use_data %>%
  group_by(Building=building, Floor=floor) %>%
  summarize(`Max computers in use`=max(num_in_use)) %>%
  inner_join(computers, by=join_by(Building==building, Floor==floor)) %>%
  mutate(`% of total`=round(`Max computers in use`/num_computers*100, 0)) %>%
  select(Building, Floor, `Max computers in use`, `% of total`)
```

### Daily maximums, by building and area

```{r message=FALSE}
#| code-fold: true
in_use_data %>%
  group_by(timestamp, building) %>%
  summarize(num_in_use=sum(num_in_use)) %>%
  group_by(building, Date=date(timestamp)) %>%
  summarize(max_in_use=max(num_in_use)) %>%
  inner_join(computers_by_building, by=join_by(building)) %>%
  mutate(`% computers in use`=max_in_use/num_computers*100) %>%
  ggplot(aes(x=Date, y=`% computers in use`)) +
    geom_line() +
    facet_wrap(~building)
```

```{r message=FALSE}
#| code-fold: true
in_use_data %>%
  group_by(area, Date=date(timestamp)) %>%
  summarize(max_in_use=max(num_in_use)) %>%
  inner_join(computers, by=join_by(area)) %>%
  mutate(`% computers in use`=max_in_use/num_computers*100) %>%
  ggplot(aes(x=Date, y=`% computers in use`)) +
    geom_line() +
    facet_wrap(~area)
```
